---
title: "Alcohol Consumption"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(dplyr)
```

The dataset below was downloaded from Gapminder. It includes the recorded and/or estimated alcohol consumption per adult (age 15+) in litres pure alcohol by country from 1985 to 2008.

```{r}
alc <- read.csv('alcohol consumption.csv')

# adjust columns and remove N/A values from alcohol column
alc_df <- gather(alc, "year", "alcohol", -1) %>% subset(!is.na(alcohol))

# change country column name
colnames(alc_df)[1] <- 'country'

# remove 'X' from years column
alc_df$year <- substring(alc_df$year, 2, 5)
```

### Data Exploration

```{r}
library(ggplot2)

qplot(x = alcohol, data = alc_df, binwidth = 0.5) +
  scale_x_continuous(breaks = seq(0,25,1))
```

The alcohol consumption data does not have a simple distribution. It appears that most of the records indicate an alcohol consumption of 10 litres or less per adult. The distribution could be loosely described as right-skewed, but there are clear peaks at both 1 litre and 10 litres, as well as some peaks and valleys in between.

```{r}
qplot(x = 1, y = alcohol, data = alc_df, geom = 'boxplot')
```

A boxplot of all of the alcohol consumption data shows a median around 6.5-7 litres per adult, with the middle 50% of data falling between 3 litres and 10 litres. There are also a couple outliers greater than 22.5 litres per adult.

```{r}
qplot(x = year, y = alcohol, data = alc_df, geom = 'boxplot')
```

The boxplots of alcohol consumption by year shows a large variation in alcohol consumption over the years. It also seems that some of the years may either have not much data points or data with very little variation.

```{r}
table(alc_df$year)

# create subset for years with at least 4 data values
subset_alc <- alc_df %>% transform(year = as.numeric(year)) %>% subset(alc_df$year %in% c(1990, 1993, 1994, 1995, 2005, 2008))

by(subset_alc$alcohol, subset_alc$year, summary)
```

By help of a table of the data, we can see that many of the years only have 1-2 data points. With such little data we cannot be confident at all about the data for these years. To attempt to deal with this issue, we have subsetted the data and only included years with at least 4 data points.

```{r}
ggplot(aes(group = year, x = year, y = alcohol), data = subset_alc) +
  geom_boxplot()
```

The boxplots for the years with at least 4 data points still shows a lot of variation. There isn't a simple trend that can be detected in this plot. Below we'll see if scatterplots can help make sense of the alcohol consumption by year.

```{r}
ggplot(aes(x = year, y = alcohol), data = subset_alc) +
  geom_point()
```

It's clear here that most of our data is in the years 2005 and 2008, but the alcohol consumption for all of the year's seems to vary quite a bit.

```{r}
alcByYear <- subset_alc %>% transform(year = as.numeric(year)) %>% group_by(year) %>%
  summarise(alcohol_mean = mean(alcohol),
            alcohol_median = median(alcohol),
            n = n())

ggplot(aes(x = year, y = alcohol_mean), data = alcByYear) +
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE)
```

Again, there is not a simple relationship that can be detected here. Likely, there are not enough data points and too much variation between countries to be able to see a relationship between alcohol and year. Below we'll try to investigate alcohol consumption over the years by individual countries.

```{r}
alcByCountry <- alc_df %>% group_by(country) %>% arrange(year)

count(alcByCountry) %>% subset(n > 10)
```

Russia and Sweden are the only two countries that have more than 10 data points for alcohol consumption. We'll look at these countries' alcohol consumption over the years.

```{r}
Rus_df <- subset(alc_df, country == 'Russia') %>% transform(year = as.numeric(year))
Swe_df <- subset(alc_df, country == 'Sweden') %>% transform(year = as.numeric(year))

Rus <- ggplot(aes(x = year, y = alcohol), data = Rus_df) +
  geom_point()

Swe <- ggplot(aes(x = year, y = alcohol), data = Swe_df) +
  geom_point()

library(gridExtra)
grid.arrange(Rus, Swe, ncol = 1)
```

Based on these two countries, which admittedly is still a small sample, it appears that alcohol consumption has been generally increasing as time goes on. In recent years it seems that alcohol consumption has been on the decline in Sweden, but in Russia it increased from 2005 to 2008. Taking the variability due to the different countries out of the analysis, we get a clearer picture of the relationship between alcohol consumption over time.

```{r}
cor.test(Rus_df$year, Rus_df$alcohol)
```
```{r}
cor.test(Swe_df$year, Swe_df$alcohol)
```

The r values further demonstrate the positive correlation between alcohol consumption and year in both Sweden and Russia.

## Further Analysis

Going forward, we could do a further analysis between the years 2005 and 2008 since this is where most of the data points came from. It would be interesting to explore whether alcohol consumption has generally increased or decreased from 2005 to 2008. In Russia it increased, but in Sweden it decreased. We could perform some data exploration prior to performing further statistical tests to get an idea of which direction alcohol consumption changed between 2005 to 2008.